---
title: "Machine Learning Course Project"
author: "Jessica Peterka-Bonetta"
date: "24. May 2015"
output: html_document
---

# Summary

# Exploratory Data Analysis
A first look at the data reveals a total number ob 160 variables, consisting of some general information (username, timestamp), a large number of variables representing different measurements and their evaluations (e.g. "roll_belt", amplitude_roll_belt" or "total_accel_arm") as well as information about the class of exercise performed. The latter consists of five different ways to perform the weight lifting exercise, where class A corresponds to the exercise beeing performed correctly and class B to E correspond to common mistakes. 

The data is divided into two sets: a training set with aprroximatively 20.000 observations and a test set with 20 observations. The test set is reserved for testing purposes and will only be used once a model has been picked. 

```{r, echo=FALSE}
setwd("~/Desktop/coursera-r-class/MachineLearning")
library(caret)
library(lattice)
library(ggplot2)
library(doMC)
library(parallel)
library(foreach)
library(iterators)
library(randomForest)
set.seed(1234)

# Download datasets
fileUrl1 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
fileUrl2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(fileUrl1, destfile = "pml_training.csv", method = "curl")
download.file(fileUrl2, destfile = "pml_test.csv", method = "curl")

# Read in datasets
training <- read.csv("pml_training.csv")
test <- read.csv("pml_test.csv")
```

 
# Data partitioning

Luckily, the training set with which the prediction model will be trained is large enough to take subsamples of it for cross validation purposes. The training set is beeing partitioned into a sub training set with 60% of the observations and a sub testing set with 40% of the observations. 
```{r}
inTrain <- createDataPartition(y = training$classe, p = 0.6, list = FALSE)
sub_training <- training[inTrain, ]
sub_testing <- training[-inTrain, ]
dim(sub_training)
dim(sub_testing)
```

 
# Data cleaning

No prediction without variance - this is true for the outcome variable and the predictor variables alike. The first data cleaning measure therefor consists of leaving out variables with variances close to zero.  
```{r}
# Identify near zero variance variables
sub_trainingNZV <- nearZeroVar(sub_training)
# Create subset of sub_training with non-nrz variables only
sub_training <- sub_training[, -sub_trainingNZV]
# Apply filter to sub_testing and test datasets
sub_testing <- sub_testing[, -sub_trainingNZV]
# Check for number of observations left
dim(sub_training) ; dim(sub_testing)
```

The next step is to check for the proportion of missing values for each columns. 
It seems variables have either very high percentage of NA (approx. 98%), which disqualifies them as predictors, or no NA at all. Therefore, variables with more than 10% of NA are filtered out of the dataset. Note that for the sub_testing dataset the percentage of NA in the sub_training dataset is used to filter out the variables with too many NA. Also, the first column "X" representing the row name is beeing removed from the data. 

```{r}
# Compute percentage of NA for each variable in the sub_training data set
apply(sub_training, 2, function(col)sum(is.na(col))/length(col))
# Filter out variables with high NA percentage
sub_training <- sub_training[lapply(sub_training, function(x) sum(is.na(x)) / length(x) ) < 0.1 ]
variables_names <- colnames(sub_training)
sub_testing <- sub_testing[variables_names]
# Remove first column X and second column user_name
sub_training_clean <- sub_training[, 3:ncol(sub_training)]
sub_testing_clean <- sub_testing[, 3:ncol(sub_training)]
```

After cleaning data as described above, 58 variables are left. 
 
 
# Model building

The random forests approach is choosen to train a prediction model with cross validation.

```{r, cache=TRUE}
registerDoMC(3)
modelFit <- train(classe ~., data = sub_training_clean, method = "rf")
modelFit
```
 
## Out of bag error

Let´s take a closer look at the out of bag error for the final model:
```{r}
modelFit$finalModel
```
The out of bag error rate for the final model is 0.15%. 
 
## Model accuracy

The final model trained with the random forests method is tested on the sub_testing dataset to get an idea of its accuracy. 
```{r}
prediction1 <- predict(modelFit, sub_testing_clean)
confusionMatrix(prediction1, sub_testing_clean$classe)
```
The accuracy is as high as 0.9991, which means that only 0.0009% of the out-of-sample cases are misclassified by the model. 
 
# Write up

First, predictions are made for the test dataset. 
```{r}
prediction2 <- predict(modelFit, test)
```

This is the code used to generate files for upload. 
```{r}
pml_write_files = function(x) {
    n = length(x)
    for (i in 1:n) {
        filename = paste0("problem_id_", i, ".txt")
        write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, 
            col.names = FALSE)
    }
}

pml_write_files(prediction2)
```